---
title: "Fourfold Displays for 'ggplot2'"
output: rmarkdown::html_vignette
bibliography: refs.bib
csl: apa.csl
vignette: >
  %\VignetteIndexEntry{Fourfold Displays for 'ggplot2'}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(ggfourfold)
library(ggplot2)
```

`ggfourfold` provides `geom_fourfold()` and `theme_fourfold()` for drawing
fourfold displays of $2 \times 2$ and $2 \times 2 \times k$ tables with
`ggplot2`. This introduction follows the treatment in
@FriendlyMeyer:2016:DDAR [Section 4.4].

## Fourfold displays

The *fourfold display* is a special case of a radial diagram, or "polar area
chart", designed for $2 \times 2$ (or $2 \times 2 \times k$) tables
[@Fienberg:75; @Friendly:94:TR217; @Friendly:94c]. The frequency $n_{ij}$ in
each cell is shown by a quarter circle whose radius is proportional to
$\sqrt{n_{ij}}$, so that its *area* is proportional to the cell count. Like a
pie chart, it uses segments of a circle to show frequencies; unlike a pie
chart, it keeps the angles of the segments constant and varies the radius.

The main purpose of the display is to show the sample odds ratio,

$$
\hat{\theta} = \frac{n_{11} / n_{12}}{n_{21} / n_{22}}
             = \frac{n_{11} \, n_{22}}{n_{12} \, n_{21}} .
$$

An association between the two variables ($\theta \neq 1$) appears as a
tendency for one pair of diagonally opposite cells to be larger than the
other pair.

* **Color and diagonal tick marks** show the **direction of association**: the relatively larger diagonal pair is drawn in blue
(more than expected under independence) with the tick marks along that diagonal, and the other pair in red. 

* **Confidence rings** around
each quadrant give a **visual test of independence**, $H_0: \theta = 1$: in a
standardized display, the rings for adjacent quadrants overlap if and only if
the observed counts are consistent with the null hypothesis.

## Example: Berkeley admissions

The classic example, `UCBAdmissions` gives applicants to the six largest graduate departments at
Berkeley in 1973, classified by admission and gender [@Bickel-etal:75].
`geom_fourfold()` currently works on data in *long form*: map the two-level variables
to `x` (drawn left to right) and `y` (drawn top to bottom), and the cell
frequencies to `weight`. Without faceting, rows for the same cell are summed,
so this pools over departments:

```{r berkeley}
ucb <- as.data.frame(UCBAdmissions)
berkeley <- ggplot(ucb, aes(x = Admit, y = Gender, weight = Freq)) +
  theme_fourfold()
```

The *unstandardized* display (`std = "ind.max"`) shows the raw frequencies,
with the largest cell drawn at full size. The sample odds ratio,
Odds(Admit | Male) / Odds(Admit | Female), is
$(1198 / 1493) / (557 / 1278) = 1.84$: male applicants were nearly twice
as likely to be admitted.

```{r berkeley-raw, fig.width = 4, fig.height = 4}
berkeley + geom_fourfold(std = "ind.max")
```

These comparisons are difficult because there are more male applicants than
female, and because more applicants were rejected than admitted. In the
unstandardized display the confidence rings have no interpretation as a test
of $H_0: \theta = 1$.

The data can be standardized to make the comparison easier while preserving
the odds ratio. With `margin = 1` the rows (here, `Gender`) are equated, so
each quadrant shows the proportion of men or women who were admitted or
rejected: 44.5% of male applicants were admitted, compared with 30.4% of
female applicants. Now the rings for each upper quadrant overlap those of
the quadrant below it if the odds ratio does not differ from 1.

```{r berkeley-rows, fig.width = 4, fig.height = 4}
berkeley + geom_fourfold(margin = 1)
```

The default, `margin = c(1, 2)`, standardizes *both* margins. The four
quadrants then align horizontally and vertically when the odds ratio is 1,
regardless of the marginal frequencies, so the association between
admission and gender can be seen without the influence of the overall
admission rate or of the numbers of men and women who applied. This fully
standardized display is usually the most useful form.

```{r berkeley-std, fig.width = 4, fig.height = 4}
berkeley + geom_fourfold()
```

The quadrants do not align, and the 95% confidence rings, computed from the
interval $1.62 \le \theta \le 2.09$, do not overlap: the odds ratio differs
significantly from 1, apparent evidence of gender bias in favour of men.

## Stratified $2 \times 2 \times k$ tables

In a $2 \times 2 \times k$ table the third variable often defines strata, and
the question is whether the association between the first two is the same
in each stratum. Make one fourfold display per stratum with
`facet_wrap()` or `facet_grid()`; standardizing the margins within each
panel allows the pattern of association to be compared across strata even
when their marginal frequencies differ.

```{r berkeley-dept, fig.width = 7, fig.height = 5}
berkeley +
  geom_fourfold() +
  facet_wrap(vars(Dept), ncol = 3, labeller = label_both)
```

Surprisingly, for five of the six departments the odds of admission are
about the same for men and women. Only Department A differs, and in the
opposite direction: there women were about 2.86 times as likely as men to
be admitted ($1 / 0.349$). The more intense colours mark the one panel whose
odds ratio differs significantly from 1. With several strata,
`geom_fourfold()` adjusts the p-values that control this emphasis for
multiple testing, by default using Holm's method via `stats::p.adjust()`;
the confidence rings themselves are not adjusted.

This reversal of the aggregate result is an example of Simpson's paradox,
and it shows why an analysis of a table collapsed over a third variable can
mislead. Men and women applied to departments in different proportions, and
women applied in greater numbers to departments with low admission rates.

Because a fourfold display keeps the angle of each cell fixed, corresponding
cells are always in the same position in every panel. An array of fourfold
displays therefore supports *controlled comparison*, against a standard with
other things held constant, much better than an array of pie charts.
Standardizing the margins adds a further standard for comparison while
preserving the odds ratio, the quantity the display is designed to show.

## References
